---
title: "R Notebook"
output: html_notebook
---

# Summary

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, cache=FALSE}
# Synapse client
library(synapseClient)

# viz packages
library(ggthemes)
library(viridis)
library(ggrepel)
library(ggbeeswarm)
library(ComplexHeatmap)
library(circlize)
library(patchwork)

# packages for general data munging, formatting
library(feather)
library(stringr)
library(forcats)
library(broom)
library(tidyverse)

# analysis packages
library(car)
library(vegan)

my_theme_bw <- function() {
    theme_bw() +
        theme(axis.title = element_text(face = "bold", size = rel(0.9)),
              legend.title = element_text(face = "bold"),
              plot.title = element_text(face = "bold"))
}

?scale_fill_distiller
ggplot2::theme_set(my_theme_bw())
scale_fill_discrete <- function(...) scale_fill_colorblind(...)
scale_fill_continuous <- function(...) scale_fill_distiller(..., 
                                                            type = "div",
                                                            palette = "RdBu")
scale_colour_discrete <- function(...) scale_color_colorblind(...)
# scale_colour_continuous <- function(...) scale_color_viridis(...)

subtype_colors <- c("red", "yellow", "green", "cyan", "blue", "magenta")
```

# Prepare data

## Retrieve and load data

### IRWG feature matrix (Synapse)

```{r}
if (!dir.exists("../data")) {
    dir.create("../data")
}
synapseLogin()
```

*NOTE:* requires SFTP authentication

```{r}
irwg_fmx_file <- synapseClient::synGet("syn11187757", 
                                       downloadLocation = "../data/irwg/")
load(irwg_fmx_file@filePath)
irwg_fmx_df <- df
rm(df)
```

### Immunomodulator gene table (Google Sheets)

List of genes accessed [here](https://docs.google.com/spreadsheets/d/1aqOXYsU1ubkbxIZI_5p8ZRootgOAT0KweMA3LvSZ7HY/edit#gid=0) and saved as a TSV at `data/Cancer Immunomodulators - TCGA PanImmune Group - Direct Relationship.tsv`:

```{r}
im_url <- "https://docs.google.com/spreadsheets/d/1aqOXYsU1ubkbxIZI_5p8ZRootgOAT0KweMA3LvSZ7HY/edit#gid=0"
googlesheets::gs_auth()
im_gs <- googlesheets::gs_url(im_url)
```

```{r}
im_df <- googlesheets::gs_read(im_gs) %>% 
    dplyr::filter(!is.na(Gene))
```

### Immunomodulator expression

```{r include=FALSE}
im_subtype_logexpr_df <- readr::read_csv("../data/intermediate/subtype_immunomodulator_expr.csv")
```


### Methylation-immunomodulator correlations (Synapse)

```{r}
im_methcorr_file <- synapseClient::synGet("syn11630210",
                                          downloadLocation = "../data/irwg/")
im_methcorr_df <- readr::read_csv(im_methcorr_file@filePath)
```


### Immunomodulator CNV data (Synapse)

```{r}
im_cnv_amp_file <- synapseClient::synGet("syn11673281",
                                          downloadLocation = "../data/irwg/")
im_cnv_amp_df <- readr::read_tsv(im_cnv_amp_file@filePath) %>% 
    purrr::set_names(c("gene_name", "C1", "C2", "C3", "C4", "C5", "C6"))
```

```{r}
im_cnv_del_file <- synapseClient::synGet("syn11673280",
                                          downloadLocation = "../data/irwg/")
im_cnv_del_df <- readr::read_tsv(im_cnv_del_file@filePath) %>% 
    purrr::set_names(c("gene_name", "C1", "C2", "C3", "C4", "C5", "C6"))
```

## Format data for analysis

```{r}
im_subtype_feature_df <- im_subtype_logexpr_df %>% 
    dplyr::left_join(im_cnv_amp_df %>% 
                         tidyr::gather(Subtype_Immune_Model_Based, amp_diff,
                                       -gene_name),
                     by = c("Subtype_Immune_Model_Based", "gene_name")) %>% 
    dplyr::left_join(im_cnv_del_df %>% 
                         tidyr::gather(Subtype_Immune_Model_Based, del_diff,
                                       -gene_name),
                     by = c("Subtype_Immune_Model_Based", "gene_name")) %>% 
    dplyr::left_join(im_methcorr_df %>% 
                         mutate(Subtype_Immune_Model_Based = str_c("C", SubType)) %>% 
                         group_by(Subtype_Immune_Model_Based, Symbol) %>% 
                         summarise(spearman = mean(spearman)),
                     by = c("Subtype_Immune_Model_Based", "gene_name" = "Symbol"))
```


```{r, warning=FALSE}
im_ordered <- im_subtype_feature_df %>% 
    select(`Super Category`, gene_name, Subtype_Immune_Model_Based, 
           x_med, spearman, amp_diff, del_diff) %>% 
    nest(-`Super Category`) %>% 
    mutate(data = map(data, function(df) {
        df %>% 
            nest(-Subtype_Immune_Model_Based) %>%
            mutate(data = map2(data, Subtype_Immune_Model_Based, function(x, y) {
                x %>%
                    gather(stat, val, -gene_name) %>%
                    mutate(stat = str_c(stat, y, sep = "_"))
            }
            )) %>%
            unnest() %>% 
            select(-Subtype_Immune_Model_Based) %>%
            spread(stat, val) %>%
            remove_rownames() %>%
            column_to_rownames("gene_name") %>%
            as.matrix() %>%
            dist() %>%
            hclust()
    })) %>% 
    mutate(data = map(data, function(hc) {
        tibble(gene_name = hc$labels[hc$order], gene_pos = seq_along(hc$labels))
    })) %>% 
    unnest() %>% 
    mutate(gene_pos = as.character(gene_pos),
           gene_pos = str_pad(gene_pos, 2, side = "left", pad = "0"),
           gene_order = str_c(`Super Category`, gene_pos, sep = "_")) %>% 
    arrange(gene_order) %>% 
    pluck("gene_name")

im_subtype_feature_df <- im_subtype_feature_df %>% 
    mutate(gene_name = factor(gene_name, levels = im_ordered),
           Subtype_Immune_Model_Based = str_replace(Subtype_Immune_Model_Based,
                                                    "C", ""))
```


## Viz

```{r, fig.asp=1.2}

heatmap_panel <- function(df, panel_scale, panel_label, scale_label) {
    df %>% 
        ggplot2::ggplot(ggplot2::aes(x = Subtype_Immune_Model_Based, 
                                     y = gene_name)) +
        ggplot2::geom_tile(aes_string(fill = panel_scale), 
                           colour = "gray60", size = 0.3) +
        ggplot2::scale_fill_distiller(scale_label, 
                                      type = "div", palette = "RdBu") +
        ggplot2::facet_grid(`Super Category` ~ Subtype_Immune_Model_Based, 
                            space = "free", scales = "free", drop = TRUE) +
        ggplot2::ylab("") +
        ggplot2::xlab(panel_label) +
        ggplot2::theme(axis.text.x = ggplot2::element_blank(),
                       axis.text.y = ggplot2::element_blank(),
                       axis.ticks.x = ggplot2::element_blank(),
                       axis.ticks.y = ggplot2::element_blank(),
                       axis.title.x = ggplot2::element_text(size = 8, angle = 45,
                                                            hjust = 1, vjust = 1),
                       strip.text.x = ggplot2::element_text(face = "bold", 
                                                            angle = 0,
                                                            hjust = 0, 
                                                            size = 8,
                                                            margin = margin(l = 1, b = 2)),
                       strip.text.y = ggplot2::element_blank(),
                       strip.background = ggplot2::element_blank(),
                       panel.spacing.y = ggplot2::unit(4, "points"),
                       panel.spacing.x = ggplot2::unit(1, "points"),
                       legend.position = "top",
                       legend.title = ggplot2::element_text(size = 8),
                       legend.text = ggplot2::element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
                       legend.key.width = ggplot2::unit(12, "points")) +
        ggplot2::guides(fill = ggplot2::guide_colorbar(title.position = "top"))
}

p_expr <- im_subtype_feature_df %>% 
    filter(`Super Category` != "Antigen presentation") %>% 
    heatmap_panel("x_med", "mRNA Expression", "Median log10(x)")
p_methcorr <- im_subtype_feature_df %>% 
    filter(`Super Category` != "Antigen presentation") %>% 
    heatmap_panel("spearman", "Expression-Methylation", "Spearman Corr.")
p_cnv_amp <- im_subtype_feature_df %>% 
    filter(`Super Category` != "Antigen presentation") %>% 
    heatmap_panel("amp_diff", "Amplification", "Freq. (O - E)")
p_cnv_del <- im_subtype_feature_df %>% 
    filter(`Super Category` != "Antigen presentation") %>% 
    heatmap_panel("del_diff", "Deletion", "Freq. (O - E)")

p_annot <- im_subtype_feature_df %>% 
    filter(`Super Category` != "Antigen presentation") %>% 
    ggplot2::ggplot(ggplot2::aes(x = 1, 
                                 y = gene_name)) +
    ggplot2::geom_tile(aes(fill = `Immune Checkpoint`), 
                       colour = "gray60", size = 0.3) +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::scale_y_discrete(position = "right") +
    scale_fill_manual(values = c("#333333", "#888888", "#E69F00")) +
    ggplot2::facet_grid(`Super Category` ~ ., 
                        switch = "both",
                        space = "free", scales = "free", drop = TRUE) +
    ggplot2::ylab("") + 
    ggplot2::xlab("") +
    ggplot2::theme(axis.text.x = ggplot2::element_blank(),
          axis.ticks.x = ggplot2::element_blank(),
          axis.text.y = ggplot2::element_text(size = 7),
          axis.ticks.y = ggplot2::element_blank(),
          strip.text.y = ggplot2::element_text(angle = 180, face = "bold", 
                                               hjust = 1, size = 8),
          strip.background = ggplot2::element_blank(),
          panel.spacing.y = ggplot2::unit(5, "points"),
          legend.position = "top",
          legend.direction = "vertical",
          legend.title = ggplot2::element_text(size = 8),
          legend.text = ggplot2::element_text(size = 7),
          legend.key.size = ggplot2::unit(10, "points")) +
    ggplot2::guides(fill = ggplot2::guide_legend(title.position = "top"))

p_all <- p_annot + p_expr + p_methcorr + p_cnv_amp + p_cnv_del + 
    patchwork::plot_layout(nrow = 1, widths = c(2, 15, 15, 15, 15))
p_all
# ggsave("../figures/im_subtype_heatmaps.pdf", plot = p_all, width = 15, height = 20, units = "cm")

```

```{r}
im_subtype_feature_df %>% 
    filter(x_med > 0) %>% 
    gather(direction, rel_freq, amp_diff, del_diff) %>% 
    ggplot(aes(y = rel_freq, x = x_med)) +
    ggplot2::geom_point(ggplot2::aes(fill = spearman), 
                        size = 2.5, alpha = 0.8, shape = 21, colour = "black", 
                        stroke = 0.2) +
    facet_grid(direction ~ Subtype_Immune_Model_Based) +
    scale_fill_viridis_c(option = "D")
```

-----