---
title: "PanCan IRWG miRNA Correlative Analysis"
author: "James A. Eddy"
date: '`r lubridate::today()`'
output:
  html_notebook:
    code_folding: hide
    fig_width: 8
    toc: yes
    toc_float: yes
---

# Summary

Exploring associations of miRNA expression levels across tumor types in TCGA with suggested correlates defined by the Immune Response Working Group:

+ Overall Leukocyte fraction  
+ Individual Relative CIBERSORT fraction  
+ Mutation Load  
+ TCR,BCR Diversity  
+ Expression of  
    + IFNG,IFN-gamma  
    + PRF1,Perforin  
    + GZMA,Granzyme A  
    + PDCD1,PD-1  
    + CD274,PD-L1  
    + PDCD1LG2,PD-L2  
    + IL10,IL-10  
    + TGFB1,TGF-beta  
    + IDO1,IDO  
    + HLA-A  


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, cache=FALSE}
# Synapse client
library(synapseClient)

# viz packages
library(ggthemes)
library(viridis)
library(ggbeeswarm)

# packages for general data munging, formatting
library(feather)
library(stringr)
library(forcats)
library(broom)
library(tidyverse)

my_theme_bw <- function() {
    theme_bw() +
        theme(axis.title = element_text(face = "bold", size = rel(0.9)),
              legend.title = element_text(face = "bold"),
              plot.title = element_text(face = "bold"))
}
ggplot2::theme_set(my_theme_bw())
scale_fill_discrete <- function(...) scale_fill_colorblind(...)
scale_fill_continuous <- function(...) scale_fill_viridis(...)
scale_colour_discrete <- function(...) scale_color_colorblind(...)
scale_colour_continuous <- function(...) scale_color_viridis(...)
```

-----

\  

# Prepare data

## Retrieve miRNA data from Synapse

```{r}
if (!dir.exists("data")) {
    dir.create("data")
}
synapseLogin()
mirna_synfolder <- "syn6171109"
mirna_synfiles <- synapseQuery(
    sprintf('select * from file where parentId=="%s"', mirna_synfolder)
)

# download files and store data/paths in new data frame
mirna_files <- mirna_synfiles %>% 
    mutate(file_data = map(file.id, function(synid) {
        synGet(synid, downloadLocation = "data/")
    }),
    file_path = map_chr(file_data, getFileLocation))
```

## Load miRNA data

```{r}
# load sample data
mirna_sample_file <- mirna_files %>% 
    filter(file.id == "syn7222010") %>% 
    .[["file_path"]]
mirna_sample_df <- read_tsv(mirna_sample_file)

# load normalized, batch-corrected expression data
mirna_norm_file <- mirna_files %>% 
    filter(file.id == "syn7201053") %>% 
    .[["file_path"]]
mirna_norm_df <- read_csv(mirna_norm_file)
```

-----

\  

# Explore miRNA data

```{r}
# convert expression df to matrix
mirna_norm_mat <- mirna_norm_df %>% 
    select(one_of(c("Genes", mirna_sample_df$id))) %>% 
    column_to_rownames("Genes") %>% 
    as.matrix()
mirna_norm_mat[1:5, 1:5]
```

```{r}
mirna_norm_mat %>% 
    rowMeans() %>% 
    tibble(avg_abundance = .) %>% 
    ggplot(aes(x = avg_abundance + 1)) +
    geom_histogram(bins = 50) +
    scale_x_log10()
```

