---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Summary

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, cache=FALSE}
# Synapse client
library(synapseClient)

# parallel computing
library(parallel)

# viz packages
library(ggthemes)
library(viridis)
library(ggrepel)
library(ggbeeswarm)
library(ComplexHeatmap)
library(circlize)
library(patchwork)

# packages for general data munging, formatting
library(feather)
library(stringr)
library(forcats)
library(broom)
library(tidyverse)

# analysis packages
library(car)
library(vegan)

my_theme_bw <- function() {
    theme_bw() +
        theme(axis.title = element_text(face = "bold", size = rel(0.9)),
              legend.title = element_text(face = "bold"),
              plot.title = element_text(face = "bold"))
}
ggplot2::theme_set(my_theme_bw())
scale_fill_discrete <- function(...) scale_fill_colorblind(...)
scale_fill_continuous <- function(...) scale_fill_distiller(..., 
                                                            type = "div",
                                                            palette = "RdBu")
scale_colour_discrete <- function(...) scale_color_colorblind(...)
# scale_colour_continuous <- function(...) scale_color_viridis(...)

subtype_colors <- c("red", "yellow", "green", "cyan", "blue", "magenta")

```

# Prepare data

## Retrieve and load data

### IRWG feature matrix (Synapse)

```{r}
if (!dir.exists("../data")) {
    dir.create("../data")
}
synapseLogin()
```

*NOTE:* requires SFTP authentication

```{r}
irwg_fmx_file <- synapseClient::synGet("syn11187757", 
                                       downloadLocation = "../data/irwg/")
load(irwg_fmx_file@filePath)
irwg_fmx_df <- df
rm(df)
```


### TCGA miRNA expression data (Synapse)

```{r}
mirna_synfolder <- "syn6171109"
mirna_synfiles <- synapseQuery(
    sprintf('select * from file where parentId=="%s"', mirna_synfolder)
)

# download files and store data/paths in new data frame
mirna_files <- mirna_synfiles %>% 
    mutate(file_data = map(file.id, function(synid) {
        synGet(synid, downloadLocation = "../data/tcga/")
    }),
    file_path = map_chr(file_data, getFileLocation))
```


### TCGA gene expression data (Synapse)

Batch effects normalized mRNA data: syn4976363
syn4976369 ("EB++AdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.tsv")
syn4976366 (" EB++GeneExpAnnotation.tsv")

mRNA normalized, batch corrected expression values for all samples are stored as a matrix in a TSV file (Synapse ID: `syn4976369`) and can be loaded with `read_tsv()`. 

```{r}
im_expr_file <- "../data/intermediate/immunomodulators_mrna-expr.feather"
im_expr_df <- feather::read_feather(im_expr_file)
```

### Immunomodulator gene table (Google Sheets)

List of genes accessed [here](https://docs.google.com/spreadsheets/d/1aqOXYsU1ubkbxIZI_5p8ZRootgOAT0KweMA3LvSZ7HY/edit#gid=0) and saved as a TSV at `data/Cancer Immunomodulators - TCGA PanImmune Group - Direct Relationship.tsv`:

```{r}
im_url <- "https://docs.google.com/spreadsheets/d/1aqOXYsU1ubkbxIZI_5p8ZRootgOAT0KweMA3LvSZ7HY/edit#gid=0"
googlesheets::gs_auth()
im_gs <- googlesheets::gs_url(im_url)
```

```{r}
im_df <- googlesheets::gs_read(im_gs) %>% 
    dplyr::filter(!is.na(Gene))
```

## Format data for analysis

### Subset/format miRNA expression data

```{r warning=FALSE}
mirna_logexpr_file <- "../data/intermediate/mirna_logexpr_data.feather"
force_read <- FALSE
if (!file.exists(mirna_logexpr_file) | force_read) {
    # load normalized, batch-corrected expression data
    mirna_norm_file <- mirna_files %>% 
        dplyr::filter(file.id == "syn7201053") %>% 
        .[["file_path"]]
    mirna_norm_df <- readr::read_csv(mirna_norm_file, progress = FALSE)
    
    portion_id_minus_analyte_regex <- "([:alnum:]+\\-){4}[0-9]+"

    mirna_logexpr_df <- mirna_norm_df %>% 
        dplyr::select(gene_name = Genes, everything(), -Correction) %>% 
        tidyr::gather(AliquotBarcode, x, -gene_name) %>% 
        dplyr::mutate(portion_id = stringr::str_extract(
            AliquotBarcode, portion_id_minus_analyte_regex
        )) %>% 
        dplyr::inner_join(im_expr_df %>% 
                              dplyr::select(
                                  ParticipantBarcode,
                                  Representative_Expression_Matrix_AliquotBarcode
                              ) %>% 
                              dplyr::mutate(portion_id = stringr::str_extract(
                                  Representative_Expression_Matrix_AliquotBarcode, 
                                  portion_id_minus_analyte_regex
                              )),
                          by = "portion_id") %>% 
        dplyr::select(ParticipantBarcode, gene_name, x) %>% 
        dplyr::mutate(log_x = log(x + 1))
    
    write_feather(mirna_logexpr_df, mirna_logexpr_file)
    rm(mirna_norm_df)
} else {
    mirna_logexpr_df <- read_feather(mirna_logexpr_file)
}
```


## Subtype-wise correlations

```{r message=FALSE, warning=FALSE}
im_mirna_corr_file <- "../results/im_mirna_subtype_correlation.feather"
force_compute <- FALSE
if (!file.exists(im_mirna_corr_file) | force_compute) {
    
    s_list <- im_expr_df %>%
        pluck("Subtype_Immune_Model_Based") %>% 
        as.character() %>% 
        unique() %>% 
        sort()
        
    corr_df_list <- purrr::map(s_list, function(s) {
        source_df <- im_expr_df %>% 
            dplyr::filter(Subtype_Immune_Model_Based == s) %>% 
            dplyr::select(ParticipantBarcode, Subtype_Immune_Model_Based,
                          dplyr::contains("RNASeq")) %>% 
            tidyr::gather(gene_name, x, 
                          -ParticipantBarcode, -Subtype_Immune_Model_Based) %>% 
            dplyr::mutate(gene_name = stringr::str_replace(gene_name, "_RNASeq", ""),
                          log_x = log(x + 1)) %>% 
            dplyr::filter(!is.na(log_x))
        
        target_df <- mirna_logexpr_df %>% 
            dplyr::filter(ParticipantBarcode %in% source_df$ParticipantBarcode,
                          !is.na(log_x))
        
        corr_df <- inner_join(source_df, target_df,
                              by = "ParticipantBarcode",
                              suffix = c("_source", "_target")) %>% 
            dplyr::group_by(gene_name_source, gene_name_target) %>% 
            dplyr::do(broom::tidy(
                cor.test(.$log_x_source, .$log_x_target, method = "spearman")
            )) %>% 
            ungroup()
        corr_df[["p.adjust"]] <- p.adjust(corr_df$p.value, method = "BH")
        return(corr_df)
    })
        
    
    im_mirna_corr_df <- bind_rows(corr_df_list, .id = "subtype")
    write_feather(im_mirna_corr_df, im_mirna_corr_file)
} else {
    im_mirna_corr_df <- read_feather(im_mirna_corr_file)
}
```

## External miRNA data

### Regulator miRs

```{r}
mirna_causal_df <- readxl::read_excel("../data/causal_TCGA_panCancer_miRNA_immuneInfiltrate_3_15_2017.xlsx") %>% 
    .[, 1:9] %>% 
    mutate(mirna_disease = str_c(`miRNA Name`, `Tumor Type`, sep = "_"))
```

### Known immune miRs

```{r}
mirna_immune_df <- readxl::read_excel("../data/Paladini_immune_miRNAs.xlsx")
```

```{r}
mirna_immune_tidy_df <- mirna_immune_df %>%
    mutate(mirna_group = str_split(MicroRNAs, ",")) %>%
    unnest(mirna_group) %>%
    mutate(mirna_group = str_trim(mirna_group, "both")) %>%
    mutate(mirna = mirna_group) %>%
    filter(!str_detect(mirna_group, "[0-9]+[a-z]{2,}")) %>% 
    mutate(mirna = str_replace(mirna, "miR-17/92 cluster", "miR-17,miR-18a,miR-19a,miR-20a,miR-19b-1,miR-92a-1"),
           mirna = str_replace(mirna, "miR-212/132 cluster", "miR-212,miR-132"),
           mirna = str_replace(mirna, "miR-10 family", "miR-10a,miR-10b,miR-99a,miR-99b,miR-100,miR-125a,miR-125b-1,miR-125b-2"),
           mirna = str_replace(mirna, "miR-221/222", "miR-221,miR-222"),
           mirna = str_replace(mirna, "miR-10a/b", "miR-10a,miR-10b"),
           mirna = str_replace(mirna, "miR-148/152", "miR-148,miR-152"),
           mirna = str_replace(mirna, "miR-17-5p/20a", "miR-17-5p,miR-20a"),
           mirna = str_replace(mirna, "miR-221/222 cluster", "miR-221,miR-222"),
           mirna = str_replace(mirna, "miR-181a/b", "miR-181a,miR-181b"),
           mirna = str_replace(mirna, "miR-15/16", "miR-15,miR-16"),
           mirna = str_replace(mirna, "miR-181 family", "miR-181a-1,miR-181a-2,miR-181b-1,miR-181b-2,miR-181c,miR-181d"),
           mirna = str_replace(mirna, "miR-130/301", "miR-130,miR-301"),
           mirna = str_replace(mirna, "miR-99a/miR-150", "miR-99a,miR-150"),
           mirna = str_replace(mirna, "Let", "let")) %>%
    mutate(mirna = str_split(mirna, ",")) %>% 
    unnest(mirna)
```

### Predicted binding targets

```{r}
mirna_target_df <- read_tsv("../data/miRDB_v5.0_prediction_result.txt", col_names = FALSE) %>% 
    set_names(c("mirna", "gene", "score")) %>% 
    filter(str_detect(mirna, "hsa"))
```

```{r}
mirna_mrna_target_df <- read_tsv("../data/synergizer.tsv", skip = 4) %>% 
    mutate(refseq_mrna = str_split(refseq_mrna, " ")) %>% 
    unnest(refseq_mrna) %>% 
    filter(refseq_mrna %in% mirna_target_df$gene) %>% 
    left_join(mirna_target_df, by = c("refseq_mrna" = "gene")) %>% 
    mutate(mirna_target = str_c(mirna, hgnc_symbol, sep = "_"))
```

### Combined evidence

```{r}
im_mirna_support_df <- im_mirna_corr_df %>%
    filter(!is.na(p.adjust), 
           p.adjust < 0.05) %>% 
    select(subtype, mirna = gene_name_target, mrna = gene_name_source,
           estimate, p.adjust) %>% 
    mutate(mirna_target = str_c(mirna, mrna, sep = "_"),
           mirna_subtype = str_c(mirna, subtype, sep = "_"),
           mirna_short = str_extract(mirna, "(?<=hsa\\-).*"),
           mirna_short_ambiguous = str_extract(mirna_short,
                                               "[:alnum:]+\\-[:alnum:]+")) %>%
    mutate(mirbase_support = (estimate < 0) & 
               mirna_target %in% mirna_mrna_target_df$mirna_target, 
           immune_support = mirna_short %in% mirna_immune_tidy_df$mirna |
               mirna_short_ambiguous %in% mirna_immune_tidy_df$mirna) %>% 
    #        strong = abs(estimate) > 0.5,
    #        immune_strong = immune_support & strong) %>% 
I

```


# Viz

```{r}
# im_mirna_support_df %>%
#     filter(mirbase_support & immune_support) %>% 
#     dplyr::left_join(im_df, by = c("mrna" = "HGNC Symbol")) %>% 
#     mutate(end3p = str_detect(mirna_short, "3p$")) %>% 
#     # filter(mirna_short_ambiguous == "miR-182") %>%
#     ggplot(aes(x = 1, y = mirna_short_ambiguous)) +
#     geom_dotplot(aes(fill = subtype, colour = end3p), dotsize = 12,
#                  binaxis = "y", stackgroups = TRUE, positions = "all") +
#     scale_colour_manual(values = c("black", "#CCCCCC")) +
#     scale_fill_manual(values = subtype_colors) +
#     xlab("") +
#     ylab("") +
#     guides(colour = FALSE,
#            fill = guide_legend(title = element_blank())) +
#     theme(axis.text.x = element_blank(),
#           axis.ticks.x = element_blank(),
#           legend.position = "top",
#           strip.text.x = element_text(face = "bold", angle = 0, hjust = 0),
#           strip.text.y = element_text(angle = 0, face = "bold", hjust = 0),
#           strip.background = element_blank(),
#           panel.spacing.y = unit(1, "points")) +
#     facet_grid(`Super Category` + mrna ~ ., scales = "free_y", space = "free_y", drop = TRUE) %>%
#     I
```

```{r, fig.asp=0.6}
p <- im_mirna_support_df %>%
    filter(mirbase_support & immune_support) %>% 
    group_by(mrna) %>%
    mutate(num_subtype_gene = n_distinct(subtype)) %>%
    group_by(mirna) %>%
    mutate(num_subtype_mirna = n_distinct(subtype),
           num_im_mirna = n_distinct(mrna)) %>%
    ungroup() %>%
    filter(num_subtype_gene > 1,
           num_subtype_mirna > 1,
           num_im_mirna > 2) %>%
    dplyr::left_join(im_df, by = c("mrna" = "HGNC Symbol")) %>% 
    mutate(end3p = str_detect(mirna_short, "3p$")) %>% 
    ggplot(aes(x = mirna_short, y = mrna)) +
    geom_tile(aes(fill = estimate),
              colour = "gray60", size = 0.3) +
    scale_fill_distiller("Spearman Corr.", type = "seq", palette = "Blues") +
    scale_colour_manual(values = c("black", "#CCCCCC")) +
    xlab("") +
    ylab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),
          axis.text.y = element_text(size = 7),
          strip.text.x = element_text(face = "bold", angle = 0, hjust = 0, 
                                      size = 8, 
                                      margin = margin(l = 2, b = 2)),
          strip.text.y = element_blank(),
          # strip.text.y = element_text(angle = 0, face = "bold", hjust = 0, size = 8),
          strip.background = element_blank(),
          panel.spacing.y = unit(4, "points"),
          panel.spacing.x = unit(2, "points"),
          legend.position = "top",
          legend.title = ggplot2::element_text(size = 8),
          legend.text = ggplot2::element_text(size = 7, angle = 45, hjust = 1, vjust = 1),
          legend.key.width = ggplot2::unit(12, "points")) +
    ggplot2::guides(fill = ggplot2::guide_colorbar(title.position = "top")) +
    facet_grid(`Super Category` ~ subtype, 
               scales = "free", space = "free", drop = TRUE)
p
# ggsave("../figures/im_subtype_mirna.pdf", plot = p, width = 12, height = 9, units = "cm")
```

```{r}
im_mirna_support_df %>%
    filter(mirbase_support & immune_support) %>% 
    group_by(mrna) %>%
    mutate(num_subtype_gene = n_distinct(subtype)) %>%
    group_by(mirna) %>%
    mutate(num_subtype_mirna = n_distinct(subtype),
           num_im_mirna = n_distinct(mrna)) %>%
    ungroup() %>%
    filter(num_subtype_gene > 0,
           num_subtype_mirna > 0,
           num_im_mirna > 2) %>%
    dplyr::left_join(im_df, by = c("mrna" = "HGNC Symbol")) %>% 
    mutate(end3p = str_detect(mirna_short, "3p$")) %>% 
    select(1:4) %>% 
    distinct(mirna) %>% 
    # ggplot(aes(x, id = id, split = y)) +
    # geom_parallel_sets(aes(fill = subtype))
    I
    # ggplot(aes(weight = estimate, axis1 = as.factor(mrna), axis2 = as.factor(mirna_short))) +
    # geom_alluvium(aes(fill = subtype), width = 1/12, na.rm = TRUE) +
    # geom_stratum(width = 1/2, fill = "black", na.rm = TRUE)
```

